---
title: "Explore and Summarize Data : Red Wine Dataset"
output:
  html_document:
    number_sections: yes
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(corrplot)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine_df <- read.csv('wineQualityReds.csv', row.names = 1)
```

# Introduction

In this report I will explore the red wine quality dataset, a dataset that
contains 1,599 red wines with 12 variables on the chemical properties of the
wine. At least 3 wine experts rated the quality of each wine, providing a rating
between 0 (very bad) and 10 (very excellent).

- fixed acidity: most acids involved with wine or fixed or nonvolatile
   (do not evaporate readily)

- volatile acidity: the amount of acetic acid in wine, which at too high of
   levels can lead to an unpleasant, vinegar taste

- citric acid: found in small quantities, citric acid can add 'freshness'
   and flavor to wines

- residual sugar: the amount of sugar remaining after fermentation stops,
   it's rare to find wines with less than 1 gram/liter and wines with greater
   than 45 grams/liter are considered sweet

- chlorides: the amount of salt in the wine

- free sulfur dioxide: the free form of SO2 exists in equilibrium between
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial
   growth and the oxidation of wine

- total sulfur dioxide: amount of free and bound forms of S02; in low
   concentrations, SO2 is mostly undetectable in wine, but at free SO2
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

- density: the density of water is close to that of water depending on the
   percent alcohol and sugar content

- pH: describes how acidic or basic a wine is on a scale from 0 (very 
   acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

- sulphates: a wine additive which can contribute to sulfur dioxide gas 
   (S02) levels, wich acts as an antimicrobial and antioxidant

- alcohol: the percent alcohol content of the wine

- quality (score between 0 and 10)

# Exploration

# Univariate Plots Section

In this section, we will first observe the structure of the dataset. Then for
each variable of the dataset we will plot an histogram to better comprehend the 
distribution of the variable and a boxplot when needed to better visualize the 
variability of the variable.

## Structure

```{r echo=FALSE}
dim(wine_df)
```

```{r echo=FALSE}
str(wine_df)
```

The red wine quality dataset contains 1599 observations and 12 variables : 11 
are numerics (based on physicochemical tests) and 1 is an ordered factor (based on sensory data). 

```{r echo=FALSE, warning=FALSE}
# function to plot an histogram
plt_hist <- function(var, binw, brks, lim) {
  if (is.null(lim)){
    return (ggplot(data = wine_df, aes_string(x = var)) +
      geom_histogram(binwidth = binw) +
      scale_x_continuous(breaks = brks))
  } else {
      return (ggplot(data = wine_df, aes_string(x = var)) +
        geom_histogram(binwidth = binw) +
        scale_x_continuous(breaks = brks, limits = lim))
  }
}
```

## Quality
```{r echo=FALSE}
table(wine_df$quality)
```

```{r echo=FALSE}
ggplot(data = wine_df, aes(x = quality)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(0, 10, 1))
```

The score is supposed to be between 0 and 10 but we see that it falls only
between 3 and 8.The distribution seems to be normally distributed, with a most
common value of 5. More than 96% of the red wine samples have a minimum quality 
of 5.

## Fixed acidity
```{r echo=FALSE}
summary(wine_df$fixed.acidity)
```

```{r echo=FALSE, warning=FALSE}
plt_hist('fixed.acidity', 0.2, 4:16, NULL)
```

While the values are distributed between 4.60 and 15.90, most of them are
between 7.10 and 9.20.

Some values (>14.50) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('fixed.acidity', 0.2, 4:16, c(4, 15))
```

The distribution is slightly right skewed so the median of 7.90 is a better
measure of the center.

## Volatile acidity
```{r echo=FALSE}
summary(wine_df$volatile.acidity)
```

```{r echo=FALSE}
plt_hist('volatile.acidity', 0.02, seq(0.1, 1.6, 0.1), NULL)
```

While the values are distributed between 0.1200 and 1.5800, most of them are
between 0.3900 and 0.6400.

Some values (>1) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('volatile.acidity', 0.02, seq(0.1, 1.6, 0.1), c(0.1, 1))
```

The distribution seemed slightly right skewed before, now it looks rather normal
with a median approximately equal to the mean of 0.5200. We can see some peaks 
at 0.42 and 0.56.

## Citric acid
```{r echo=FALSE}
summary(wine_df$citric.acid)
```

```{r echo=FALSE}
plt_hist('citric.acid', 0.02, seq(0, 1, 0.1), NULL)
```

While the values are distributed between 0 and 1, most of them are
between 0.090 and 0.420

One value (= 1) seems to be an outlier, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('citric.acid', 0.02, seq(0, 1, 0.1), c(-0.1, 0.8))
```

The distribution seems slightly right skewed so the median of 0.260 is a better
measure of the center. We can see multiple peaks at 0, 0.25 and 0.47.

## Residual sugar
```{r echo=FALSE}
summary(wine_df$residual.sugar)
```

```{r echo=FALSE}
plt_hist('residual.sugar', 0.2, seq(0, 16, 1), NULL)
```

While the values are distributed between 0.900 and 15.500, most of them are
between 1.900 and 2.600.

Some values (>6.9) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('residual.sugar', 0.2, seq(0, 16, 1), c(0.900, 6.9))
```

The distribution looks normal around the peak but is slightly right skewed so
the median of 2.200 is a better measure of the center.

## Chlorides
```{r echo=FALSE, warning=FALSE}
summary(wine_df$chlorides)
```

```{r echo=FALSE}
plt_hist('chlorides', 0.02, seq(0, 0.7, 0.1), NULL)
```

While the values are distributed between 0.01200 and 0.61100, most of them are
between 0.07000 and 0.09000.

Some values (> 0.3) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('chlorides', 0.01, seq(0, 0.3, 0.05), c(0, 0.3))
```

The distribution looks normal around the peak but is slightly right skewed so
the median of 0.07900 is a better measure of the center.

## Free sulfur dioxide

```{r echo=FALSE, warning=FALSE}
summary(wine_df$free.sulfur.dioxide)
```

```{r echo=FALSE}
plt_hist('free.sulfur.dioxide', 2, seq(0, 72, 5), NULL)
```

While the values are distributed between 1 and 72, most of them are
between 7 and 21.

Some values (> 58) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('free.sulfur.dioxide', 2, seq(1, 72, 5), c(0, 58))
```

The distribution is right skewed so the median of 14 is a better
measure of the center.

## Total sulfur dioxide

```{r echo=FALSE, warning=FALSE}
summary(wine_df$total.sulfur.dioxide)
```

```{r echo=FALSE}
plt_hist('total.sulfur.dioxide', 5, seq(0, 290, 25), NULL)
```

While the values are distributed between 6.00 and 289.00, most of them are
between 22.00 and 62.00.

Some values (> 175) seem to be outliers, we might want to adjust the axes.


```{r echo=FALSE, warning=FALSE}
plt_hist('total.sulfur.dioxide', 5, seq(0, 175, 25), c(0, 175))
```

The distribution is slightly right skewed so the median of 38.00 is a better
measure of the center.

## Density

```{r echo=FALSE, warning=FALSE}
summary(wine_df$density)
```

```{r echo=FALSE}
plt_hist('density', 0.0005, seq(0.9900, 1.0040, 0.0025), NULL)
```

While the values are distributed between 0.9901 and 1.0037, most of them are
between 0.9956 and 0.9978.

The distribution seems normally distributed with a mean of 0.9967.

## PH

```{r echo=FALSE, warning=FALSE}
summary(wine_df$pH)
```

```{r echo=FALSE}
plt_hist('pH', 0.1, seq(2.7, 4.1, 0.1), NULL)
```

While the values are distributed between 2.740 and 4.010, most of them are
between 3.210 and 3.400.

The distribution seems normally distributed with a mean of 3.311.

## Sulphates

```{r echo=FALSE, warning=FALSE}
summary(wine_df$sulphates)
```

```{r echo=FALSE}
plt_hist('sulphates', 0.1, seq(0.3, 2, 0.1), NULL)
```

While the values are distributed between 0.3300 and 2.0000, most of them are
between 0.5500 and 0.7300.

Some values (> 1.5) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('sulphates', 0.1, seq(0.3, 1.5, 0.1), c(0.3, 1.5))
```

The distribution is slightly right skewed so the median of 0.6200 is a better
measure of the center.

## Alcohol
```{r echo=FALSE, warning=FALSE}
summary(wine_df$alcohol)
```

```{r echo=FALSE}
plt_hist('alcohol', 0.1, seq(8, 15, 0.5), NULL)
```

While the values are distributed between 8.40 and 14.90, most of them are
between 9.50 and 11.10.

Some values (> 14) seem to be outliers, we might want to adjust the axes.

```{r echo=FALSE, warning=FALSE}
plt_hist('alcohol', 0.1, seq(8, 14, 0.5), c(8, 14))
```

The distribution is slightly right skewed so the median of 10.20 is a better
measure of the center.

# Univariate Analysis

__What is the structure of your dataset?__

The red wine quality dataset contains 1599 observations and 12 variables : 11 
are numerics (based on physicochemical tests) and 1 is an ordered factor (based
on sensory data).

What I found :

- the mode for the quality is 5 and more than 96% of the red wine samples have a
minimum quality of 5.
- the distribution of the fixed acidity is right skewed and the median is 7.90
- the distribution of the volatile acidity is normally distributed and the mean 
is 0.5200
- the distribution of the citric acid is right skewed, the median is 0.260 and 
132 observations have a citric acid of 0.
- the distribution of the residual sugar is right skewed and the median is 2.200
- the distribution of the chlorides is slightly right skewed and the median is 
0.07900
- the distribution of the free sulfur dioxide is right skewed and the median is
14
- the distribution of the total sulfur dioxide is right skewed and the median is
38
- the distribution of the density is normally distributed and the mean is 0.9967
- the distribution of the pH is normally distributed and the mean is 3.311
- the distribution of the sulphates is right skewed and the median is 0.6200
- the distribution of the alcohol percentage is right skewed and the median is 
10.20

__What is/are the main feature(s) of interest in your dataset?__

The main feature of interest in our dataset is the quality. A good question to 
ask ourself would be to know which variables contribute to a high quality wine.

__What other features in the dataset do you think will help support your__
__investigation into your feature(s) of interest?__

So far I can't really put aside any variables so I would say that all the other 
11 variables can at this stage support my investigation into my feature of
interest.

__Did you create any new variables from existing variables in the dataset?__

I didn't create a new variable.

__Of the features you investigated, were there any unusual distributions?__
__Did you perform any operations on the data to tidy, adjust, or change the__
__form of the data? If so, why did you do this?__

Some distributions caught my attention. First, the quality is supposed to be 
between 0 and 10 but the range of the dataset's ratings is only from 3 to 8 
supposing that the extreme ratings are very rare or even impossible. Then, in 
the citric acid distribution there are 132 observations with a citric acid value
of 0. Even though it is stated that it is found in same quantities, it is still 
8% of the dataset without citric acid.

The dataset was already tidy and there was no missing values so I did not have 
to perform any action during the exploration. However, some outliers are present
in some distributions so I have to take that into consideration for my further 
explorations.

# Bivariate Plots Section

## Correlation Matrix

Let's look at a correlation matrix to try to understand the relationship between
variables.

```{r echo=FALSE, warning=FALSE}
M <- cor(wine_df)
corrplot(M, method='circle', tl.col = 'black', tl.srt=45)
```

```{r echo=FALSE, warning=FALSE}
cor(wine_df)
```

For a correlation coefficient r, we define :

- Strong relationship : 0.7 <= |r| <= 1.0
- Moderate relationship : 0.3 <= |r| < 0.7
- Weak relationship : 0.0 <= |r| < 0.3

We assume that we are only interested in the relationships between all variables
that are at least moderate, and for the relationships involving our main feature
(quality) the ones that have a correlation coefficient of at least +/-0.2.

Following the previous statement, we notice the following relationships :

|relationship|correlation coefficient|Strength|Direction|
|:----------:|:---------------------:|:------:|:-------:|
| citric acid - fixed acidity | 0.67170343 | moderate | positive |
| citric acid - volatile acidity | -0.552495685 | moderate | negative |
| total sulfure - free sulfure | 0.667666450 | moderate | positive |
| density - fixed acidity | 0.66804729 | moderate | positive |
| density - citric acid | 0.36494718 | moderate | positive |
| density - residual sugar | 0.355283371 | moderate | positive |
| ph - fixed acidity | -0.68297819 | moderate | negative |
| ph - citric acid | -0.54190414 | moderate | negative |
| sulphates - citric acide | 0.31277004 | moderate | positive |
| sulphates - chlorides  | 0.371260481 | moderate | positive |
| alcohol - density | -0.49617977 | moderate | negative |
| quality - volatile acidity | -0.390557780 | moderate | negative |
| quality - alcohol | 0.47616632 | moderate | positive |
| quality - citric acid | 0.22637251 | weak | positive |
| quality - sulphates | 0.251397079 | weak | positive |

We will see these relationships more in details in the following sections.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# when the 2 variables are continuous
plot_bivar <- function(v1, v2) {
  return (ggplot(data = wine_df, aes_string(x = v1, y = v2)) +
            geom_point(alpha = 1/5))}
```

## citric acid - fixed acidity

```{r echo=FALSE, warning=FALSE}
plot_bivar('citric.acid', 'fixed.acidity')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(citric.acid) %>%
  summarise(fixed.acidity_mean = mean(fixed.acidity)) %>%
  arrange(citric.acid)
ggplot(data = tmp, aes(x = citric.acid, y = fixed.acidity_mean)) +
  geom_line()
```

It appears that the more citric.acid there is, the more fixed.acidity there is.
However, there are a lot of variations when the value of citric.acid increases.

## citric acid - volatile acidity

```{r echo=FALSE, warning=FALSE}
plot_bivar('citric.acid', 'volatile.acidity')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(citric.acid) %>%
  summarise(volatile.acidity_mean = mean(volatile.acidity)) %>%
  arrange(citric.acid)
ggplot(data = tmp, aes(x = citric.acid, y = volatile.acidity_mean)) +
  geom_line()
```

It appears that the more citric.acid there is, the less volatile.acidity there
is. However, there are still a lot of variations.

## total sulfure - free sulfure

```{r echo=FALSE, warning=FALSE}
plot_bivar('free.sulfur.dioxide', 'total.sulfur.dioxide')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(free.sulfur.dioxide) %>%
  summarise(total.sulfur.dioxide_mean = mean(total.sulfur.dioxide)) %>%
  arrange(free.sulfur.dioxide)
ggplot(data = tmp, aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide_mean)) +
  geom_line()
```

It appears that the more free.sulfur.dioxide there is, the more 
total.sulfur.dioxide there is. There is a peak around around 37 of 
free.sulfur.dioxide.

## density - fixed acidity

```{r echo=FALSE, warning=FALSE}
plot_bivar('density', 'fixed.acidity')
```


```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(density) %>%
  summarise(fixed.acidity_mean = mean(fixed.acidity)) %>%
  arrange(density)
ggplot(data = tmp, aes(x = density, y = fixed.acidity_mean)) +
  geom_line()
```

It appears that the more density there is, the more fixed.acidity there is.
However there are a lot of variations.

## density - citric acid

```{r echo=FALSE, warning=FALSE}
plot_bivar('density', 'citric.acid')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(density) %>%
  summarise(citric.acid_mean = mean(citric.acid)) %>%
  arrange(density)
ggplot(data = tmp, aes(x = density, y = citric.acid_mean)) +
  geom_line()
```

It appears that the more density there is, the more citric.acid there is.
However there are a lot of variations.


## density - residual sugar

```{r echo=FALSE, warning=FALSE}
plot_bivar('density', 'residual.sugar')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(density) %>%
  summarise(residual.sugar_mean = mean(residual.sugar)) %>%
  arrange(density)
ggplot(data = tmp, aes(x = density, y = residual.sugar_mean)) +
  geom_line()
```

The relationship looks quite weak, even though there are so peaks we can't be
sure there is a real relationship between these two variables.

## ph - fixed acidity

```{r echo=FALSE, warning=FALSE}
plot_bivar('pH', 'fixed.acidity')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(pH) %>%
  summarise(fixed.acidityr_mean = mean(fixed.acidity)) %>%
  arrange(pH)
ggplot(data = tmp, aes(x = pH, y = fixed.acidityr_mean)) +
  geom_line()
```

It appears that the more pH there is, the less fixed.acidity there is. It is
indeed logical as higher values of pH correspond to more basic liquid.

## ph - citric acid

```{r echo=FALSE, warning=FALSE}
plot_bivar('pH', 'citric.acid')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(pH) %>%
  summarise(citric.acid_mean = mean(citric.acid)) %>%
  arrange(pH)
ggplot(data = tmp, aes(x = pH, y = citric.acid_mean)) +
  geom_line()
```

It appears that the more pH there is, the less citric acid there is. It is
indeed logical as higher values of pH correspond to more basic liquid.

## sulphates - citric acide

```{r echo=FALSE, warning=FALSE}
plot_bivar('citric.acid', 'sulphates')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(citric.acid) %>%
  summarise(sulphates_mean = mean(sulphates)) %>%
  arrange(citric.acid)
ggplot(data = tmp, aes(x = citric.acid, y = sulphates_mean)) +
  geom_line()
```

It appears that the more citric.acid there is, the more sulphates there is, 
especially after 0.75 of citric.acid where the amount of sulphates increases
a lot.

## sulphates - chlorides

```{r echo=FALSE, warning=FALSE}
plot_bivar('sulphates', 'chlorides')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(sulphates) %>%
  summarise(chlorides_mean = mean(chlorides)) %>%
  arrange(sulphates)
ggplot(data = tmp, aes(x = sulphates, y = chlorides_mean)) +
  geom_line()
```

The relationship between sulphates and chlorides is not quite clear. There are
some peaks but it seems quite random.

## alcohol - density

```{r echo=FALSE, warning=FALSE}
plot_bivar('alcohol', 'density')
```

```{r echo=FALSE, warning=FALSE}
tmp <- wine_df %>%
  group_by(alcohol) %>%
  summarise(density_mean = mean(density)) %>%
  arrange(alcohol)
ggplot(data = tmp, aes(x = alcohol, y = density_mean)) +
  geom_line()
```

It appears that the more alcohol there is, the less density there is.

## quality - volatile acidity

```{r echo=FALSE, warning=FALSE}
ggplot(data = wine_df, aes(x = as.factor(quality), y = volatile.acidity)) +
  geom_boxplot()
```

We can observe a trend right here : it seems that lower volatile.acidity mean
higher quality.

## quality - alcohol

```{r echo=FALSE, warning=FALSE}
ggplot(data = wine_df, aes(x = as.factor(quality), y = alcohol)) +
  geom_boxplot()
```

Apart from the value for the quality 5, we can observe a trend right here : it 
seems that higer alcohol mean higher quality.

## quality - citric acid

```{r echo=FALSE, warning=FALSE}
ggplot(data = wine_df, aes(x = as.factor(quality), y = citric.acid)) +
  geom_boxplot()
```

We can observe a trend right here : it seems that higher citric.acid mean
higher quality.

## quality - sulphates

```{r echo=FALSE, warning=FALSE}
ggplot(data = wine_df, aes(x = as.factor(quality), y = sulphates)) +
  geom_boxplot()
```

We can also observe a trend right here : it seems that higher sulphates mean
higher quality.

# Bivariate Analysis

__Talk about some of the relationships you observed in this part of the__
__investigation. How did the feature(s) of interest vary with other features__
__in the dataset?__

I started my analysis by creating a correlation matrix in order to understand 
better the relationship between variables.

I narrowed the number of relationships that I was interested in by only keeping 
the  relationships between all variables that are at least moderate, and for the
relationships involving our main feature (quality) the ones that have a 
correlation coefficient of at least +/-0.2. I did that so I could focus only on 
the most predominent relationships.

Of the 15 relationships kept for exploration, 4 concerned the quality variable
and I observed these trends :

- less volatile acidity means higher quality : it makes sense as a high level of
volatile acidity can lead to an unpleasant vinegar taste.
- more alcohol means higher quality : the percentage of alcohol is probably 
responsible for the taste.
- more citric acid means higher quality : it makes sense as citric acid can add 
freshness and flavor to wines. 
- more sulphates means higher quality.

__Did you observe any interesting relationships between the other features__
__(not the main feature(s) of interest)?__

Of the 15 relationships kept for exploration, 11 concerned the other features 
and I observed these trends :

- more citric acid means more fixed acidity, more sulphates and less volatile 
acidity.
- more free sulfure dioxide means more total sulfure dioxide.
- more density means more fixed acidity and citric acid .
- more ph means less fixed acidity and citric acid.
- more alcohol means less density.

__What was the strongest relationship you found?__

Concerning the feature of interest,  the strongest relationship that I found was
between the quality and the alcohol with a correlation coefficient of 
0.47616632, meaning it is a moderate positive relationship.

Concerning the other features, the strongest relationship that I found was 
between the ph and the fixed acidity with a correlation coefficient of 
-0.68297819, meaning it is a moderate (almost strong) negative relationship. 
It is indeed logical as higher values of pH correspond to more basic liquid.

# Multivariate Plots Section

First I will try to visualize relationships between the feature of interest and 
2 other features, then I will try to visualize relationships between 3 other 
features.

```{r echo=FALSE, warning=FALSE}
# plot a multivar plot with the quality variable as color
plot_multivar_quality <- function(var1, var2) {
  wine_df$quality <- ordered(wine_df$quality)
  return (ggplot(data = wine_df, aes_string(x = as.name(var1), 
              y = as.name(var2), color = 'quality')) +
              geom_point(alpha = 1/2) +
              scale_color_brewer(type = 'div', palette = 3))
}
```

```{r echo=FALSE, warning=FALSE}
# divide/cut a variable by quartile
get_quartiles <- function(var) {
  return (cut(wine_df[[var]], breaks = quantile(wine_df[[var]]), 
              include.lowest = TRUE))
}
```


```{r echo=FALSE, warning=FALSE}
# plot a multivar plot with var1 as the x axis, var2 as the y axis and var3 as 
# the color
plot_multivar_3 <- function(var1, var2, var3) {
  return (ggplot(data = wine_df, aes_string(x = var1, y = var2, color = var3)) +
    geom_point(alpha = 1/2) +
      scale_color_brewer(type = 'div', palette = 3))
}
```

```{r}
# create a quality_rating variable that classify the quality in 3 categories
wine_df$quality_rating <- ifelse(wine_df$quality < 5, 'Bad', 
            ifelse(wine_df$quality < 7, 
            'Average', 'Good'))
wine_df$quality_rating <- ordered(wine_df$quality_rating, 
            levels = c('Bad', 'Average', 'Good'))
```

In this new section, I created a new variable *quality_rating* containing the 
rating of the wine (Bad, Average or Good) according to the quality so I could 
facet wrap any future visualization with that variable.


## Quality features

### Quality - volatile acidity - alcohol

```{r echo=FALSE}
plot_multivar_quality('alcohol', 'volatile.acidity')
plot_multivar_quality('alcohol', 'volatile.acidity') + 
      facet_wrap(~wine_df$quality_rating)
```

High quality wines seems to have high alcohol and low volatile acidity.


### Quality - volatile acidity - citric acid

```{r echo=FALSE}
plot_multivar_quality('volatile.acidity', 'citric.acid')
plot_multivar_quality('volatile.acidity', 'citric.acid') + 
      facet_wrap(~wine_df$quality_rating)
```

High quality wines seems to have low volatile acidity and high citric acid.


### Quality - volatile acidity - sulphates

```{r echo=FALSE}
plot_multivar_quality('sulphates', 'volatile.acidity')
plot_multivar_quality('sulphates', 'volatile.acidity') + 
      facet_wrap(~wine_df$quality_rating)
```

High quality wines seems to have high sulphates and low volatile acidity.

### Quality - alcohol - citric acid

```{r echo=FALSE}
plot_multivar_quality('alcohol', 'citric.acid')
plot_multivar_quality('alcohol', 'citric.acid') + 
      facet_wrap(~wine_df$quality_rating)
```

We can see that high alcohol tends to high quality wines but we can't really say 
anything about the citric acid here.

### Quality - alcohol - sulphates

```{r echo=FALSE}
plot_multivar_quality('alcohol', 'sulphates')
plot_multivar_quality('alcohol', 'sulphates') + 
      facet_wrap(~wine_df$quality_rating)
```

High quality wines seems to have high alcohol and high sulphates.

### Quality - citric acid - sulphates

```{r echo=FALSE}
plot_multivar_quality('sulphates', 'citric.acid')
plot_multivar_quality('sulphates', 'citric.acid') + 
      facet_wrap(~wine_df$quality_rating)
```

We can see that high sulphates tends to high quality wines but we can't really
say anything about the citric acid here.

### Quality - volatile acidity - chloride

```{r echo=FALSE, warning=FALSE}
plot_multivar_quality('chlorides', 'volatile.acidity') +
  scale_x_continuous(limits = c(0, 0.2))
plot_multivar_quality('chlorides', 'volatile.acidity') +
  scale_x_continuous(limits = c(0, 0.2)) + 
      facet_wrap(~wine_df$quality_rating)
```

High quality wines seems to have low chlorides and low volatile acidity.

### Quality - volatile acidity - density

```{r echo=FALSE, warning=FALSE}
plot_multivar_quality('volatile.acidity', 'density')
plot_multivar_quality('volatile.acidity', 'density') + 
      facet_wrap(~wine_df$quality_rating)
```

High quality wines seems to have low volatile acidity and low density (even 
though the relationship seems to be week for the density).

### Quality - alcohol - residual sugar

```{r echo=FALSE, warning=FALSE}
plot_multivar_quality('alcohol', 'residual.sugar')
plot_multivar_quality('alcohol', 'residual.sugar') + 
      facet_wrap(~wine_df$quality_rating)
```

There doesn't seem to be any meaningful relationship between  the alcohol
and residal sugar.

## Other features

### Citric - fixed acidity - volatile acidity

```{r echo=FALSE, warning=FALSE}
citric_quartile <- get_quartiles('citric.acid')
plot_multivar_3('fixed.acidity', 'volatile.acidity', 'citric_quartile')
```

High fixed acidity and low volatile acidity tends to high citric acid.

### Density - fixed acidity - citric acid

```{r echo=FALSE, warning=FALSE}
density_quartile <- get_quartiles('density')
plot_multivar_3('citric.acid', 'fixed.acidity', 'density_quartile')
```

High citric acid and high fixed acid tends to highest density

### PH - fixed acidity - citric acid

```{r echo=FALSE, warning=FALSE}
pH_quartile <- get_quartiles('pH')
plot_multivar_3('citric.acid', 'fixed.acidity', 'pH_quartile')
```

Low citric acid and low fixed acidity tends to high pH.

# Multivariate Analysis

__Talk about some of the relationships you observed in this part of the__
__investigation. Were there features that strengthened each other in terms of__
__looking at your feature(s) of interest?__


The following combinations seems to contribute to a high quality wine :

- high alcohol with low volatile acidity
- high alcohol with high sulphates

These were the relationships that were the easiest to find.


__Were there any interesting or surprising interactions between features?__

The thing that surprised me the most is that there is no meaningful relationship
between residual sugar and alcohol. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = wine_df, aes(x = quality)) +
  geom_bar(aes(y = (..count..)/sum(..count..)), fill="#42c8f4", alpha=0.50) +
  geom_text(aes(y = ((..count..)/sum(..count..)), 
            label = scales::percent((..count..)/sum(..count..))), 
            stat = "count", 
            vjust = -0.25) +
  labs(title="Distribution of Quality", x="Quality", y="# of wines") +
  scale_x_continuous(breaks = seq(0, 10, 1))
  
```

### Description One

The plot above is a bar plot showing the distribution of the quality (from 0 to
10) in the red wines dataset.

The quality is supposed to be between 0 and 10 but we see that it falls only
between 3 and 8. Maybe no such things as really good wines or really bad wines 
exist, or maybe the dataset doesn't have these wines. Additionally, more 
than 96% of the red wine samples have a minimum quality of 5, meaning there are
not a lot of bad wines in the dataset.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data = wine_df, aes(x = as.factor(quality), y = alcohol)) +
  geom_boxplot(fill="#42c8f4", alpha=0.50) +
  labs(title="Box Plots of Alcohol by Quality", x="Quality", 
       y="Alcohol Percentage")
```

This plot shows a box plot of the alcohol percentage for each quality. This is a
good way to represent the relationship between the alcohol and the quality as it
allows to see the evolution of the mean and the variablity of the alcohol 
percentage for each quality.

We notice that apart from  the boxplot for quality 5, a trend seems to be
emerging : it seems that a higer percentage of alcohol tends to a higher 
quality.

### Description Two

Apart from the value for the quality 5, we can observe a trend right here : it 
seems that higher alcohol percentage leads to higher quality.


### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Three}
wine_df$quality <- ordered(wine_df$quality)
ggplot(data = wine_df, aes_string(x = 'alcohol', y = 'volatile.acidity', 
    color = 'quality')) +
  geom_point(alpha = 1/2) +
  scale_x_continuous(breaks = seq(8, 15, 0.5), limits = c(8, 14)) +
  scale_y_continuous(breaks = seq(0.1, 1.6, 0.1), limits = c(0.1, 1.2)) +
  scale_color_brewer(type = 'div', palette = 5) +
  labs(title="Quality vs Alcohol and Volatile Acidity", x="Alcohol Percentage",
       y="Volatile Acidity")
```

### Description Three

This scatterplot shows the relationship between the alcohol percentage and the 
volatile acidity, while showing at the same time the quality of each 
observation.

We notice two clusters of points :

- a cluster of low quality wines (red-orange) at the top left
- a cluster of high quality wines (blue) at the bottom right.

We can then state that high quality red wines tend to have high alcohol 
percentage (as seen previously) but also low volatile acidity.

This observation is not surprising as a too high level of volatile acidity can
lead to an unpleasant, vinegar taste, leading to a lower quality.

------

# Reflection

This project was interesting because it allowed me to put into practice the 
different steps of Exploratory Data Analysis with a powerful language like R.

The dataset I worked on is the red wine quality dataset. This dataset contains 
1599 observations and 12 variables : 11 are numerics (based on physicochemical 
tests) and 1 is an ordered factor (based on sensory data).

First of all, I did an univariate exploration. First, I observed the
structure of the dataset by displaying its dimensions, and the types of its
variables. Then, for each of the variables in the dataset, I displayed a summary
and its histogram to get an overview of its distribution. This allowed me to
know how it was distributed (right skewed, left skewed or normal) and if there 
were any outliers. It also allowed me to strengthen my understanding of the
dataset. After this exploration, I chose to focus mainly on the quality feature,
and I asked myself what were the variables contributing to a high quality wine.

After that, I did a bivariate exploration. I started my analysis by creating a 
correlation matrix in order to understand better the relationship between the
variables. I narrowed the number of relationships that I was interested in by 
only keeping the relationships between all variables that are at least moderate,
and for the relationships involving the main feature (quality) the ones that 
have a correlation coefficient of at least +/-0.2. I did that so I could focus 
only on the most predominent relationships. Concerning the main feature, I ended
up with 4 relationships : less volatile acidity means higher quality, more 
alcohol means higher quality, more citric acid means higher quality and more 
sulphates means higher quality.

On the final part of the EDA I did a multivariate exploration. Since there were 
many variables to consider and many variable associations that could be made, 
I first decided to focus on the relationships involving the variable of interest
(quality) and 2 other variables.  I could for instance understand that high 
quality wines seems to have high alcohol, low volatile acidity (responsible 
of vinegar taste at high quantity so it is logical) and high sulphates. Then, I
focused myself of a few set of 3 other variables that showed correlations in the
bivariate exploration. I could for instance understand that low acid citric and 
low fixed acidity tends to high pH, which is logical as higher values of pH 
correspond to more basic (less acid) solutions.

During this project, I encountered difficulties mainly when interpreting the 
plots of the Multivariates Exploration. Indeed, when a third variable is added,
the plot sometimes immediately becomes less clear and the relationships much 
less obvious to determine. To counter this problem, I have created a new 
variable quality_rating which classifies the quality variable into 3 categories 
"bad", "average" and "good". I then plotted the variables again but made a 
facet_wrap with this new quality_rating variable. This allowed me to focus only
on wines with good quality_rating, and it helped me to discover some trends.

Among the successes of this project, I was especially surprised at how much the
correlation matrix helped me in this exploration. It allowed me to guide my 
analysis and discover patterns and trends. This is definitely something I will 
rely on in my future analyses.

In the future, the analysis could be enriched by combining the red wine quality 
dataset with the white wine quality dataset. It might be interesting to 
determine the commonalities and differences between these two datasets, and this
could also allow us to discover new insights about the 2 datasets.

# Resources

https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html

https://minimaxir.com/2015/02/ggplot-tutorial/